---
title: "Discussion 4"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Consider a large population which has true mean $\mu$ and true standard deviation $\sigma$. We take a sample of size 3 from this population, thinking of the sample as the RVs $X_1, X_2, X_3$ where $X_i$  can be considered iid (independent identically distributed). We are interested in estimating $\mu$.

    a. Consider the estimator $\hat{\mu_1}=X_1 +X_2 -X_3$. What is the mean of this estimator? 
  
    b. Find the variance of $\hat{\mu_1}$. 
  
    c. Consider the estimator $\hat{\mu_2}=\frac{X_1 + X_2 + X_3}{3}$. What is the mean of this estimator?
  
    d. Find the variance of $\hat{\mu_2}$.
  
    e. Now, consider the estimator $\hat{\mu_3}=\frac{X_1 +2X_2 +3X_3}{6}$. What is the mean of this estimator?
  
    f. Find the variance of $\hat{\mu_3}$. 
  
    g. Which of these three estimators is preferable? Why?
  
    <!-- h. Now let $\mu=100$ and $\sigma=10$. Define the estimator $\hat{\mu_4}=\frac{X_1 +X_2 +X_3}{3}+\epsilon X_1$. Compute the Mean Squared Error: $\text{MSE} = \text{Bias}(\hat{\mu_4}) + \text{Var}(\hat{\mu_4})$. -->

    <!-- g. Is  $\hat{\mu_4}$ ever a better estimator than  $\hat{\mu_1}$? If so, find a value of $\epsilon$ for which this is the case. Is $\hat{\mu_4}$ ever better than $\hat{\mu_2}$? Than $\hat{\mu_3}$? -->

2. A packing plant fills bags with cement. The weight X kg of a bag of cement can be modeled by a normal distribution with mean 50kg and standard deviation 0.7kg.
	
  	a. Find $P(X > 51)$
  	
  	b. Three bags are selected randomly. Find the probability that at least two weigh more than 51kg.
  	

3. Weights of female cats are well approximated by a normal distribution with mean 4.1kg and standard deviation of 0.6kg $X \sim N(4.1, 0.6^2)$.

    a. What proportion of female cats have weights between 3.7 and 4.4kg?
  
    b. A certain female cat has a weight that is 0.5 standard deviations above the mean. What proportion of female cats are heavier than this one?
  
    c. How heavy is a female cat whose weight is on the 80th percentile?
  
    d. A female cat is chosen at random. What is the probability that she weighs more than 4.5 kg?
  
    <!-- e. Ten female cats are chosen at random from a large population. What is the probablity that exactly 2 of them weigh more than 4.5 kg? Assume that the weights of the 10 cats are independent (large population size). -->

    <!-- f. Ten female cats are chosen at random from a large population. What is the probablity that their average weight is more than 4.2 kg? Assume that the weights of the 10 cats are independent (large population size). -->

    <!-- g. Suppose 1 female cat is selected at random from the given distribution and 1 male cat of that same breed is also selected at random. Male weights are well approximated by a normal distribution with mean of 4.3 and standard deviation of 0.2; $Y~\sim(4.3, 0.2^2)$. Define a new random variable: $D=Y-X$. Describe the distribution of $D$ and calculate the probability that $D>0$ assuming X and Y are independent. What does this value mean in the context of the problem? -->

4. Take a look at the `R`-code below. [**Note**: You can also find this as a separate .Rmd file on Canvas. Everything after a pound symbol (\#) is a comment, and NOT part of the `R` code.]

    a. Walk through it, one line at a time, and try to make sense of it. 
    
    b. Run all the code, **EXCEPT FOR THE LAST LINE**. This should generate a bunch of QQ-plots. Half of them are from a normal distribution, the other half an exponential distribution. Can you tell which correspond to samples from the normal distribution, and which correspond to samples from the exponential distribution?
    
    c. Repeat with `sample_size <- 100`. Does this change anything?

```{r eval = FALSE}
library(tidyverse)
library(distributions3)

## Create two random variables -- one Normal and one Exponential
X <- Normal(mu = 1, sigma = 1)
Y <- Exponential(rate = 1)

sample_size <- 10

## Create a vector of i's to use for scrambling
is <- sample(1:16)

normal_samples <- data.frame(i = is[1:8]) %>% 
  mutate(distribution = "Normal",                  # Create column that just says "Normal". 
                                                   # For book keeping.
         sample = map(i, random,                   # map runs the function "random" with
                      d = X, n = sample_size)) %>% # arguments d = X, and n = sample_size 
                                                   # for each i. I.e. it creates 8 random 
                                                   # samples of size sample_size from the 
                                                   # distribution X.
                                                   # The result is this weird list thing...
  unnest_longer(col = sample) # This get's rid of the list and simply gives us a 
                              # data.frame that we can work with...

exponential_samples <- data.frame(i = is[9:16]) %>% 
  mutate(distribution = "Exponential", 
         sample = map(i, random, d = Y, n = sample_size)) %>% 
  unnest_longer(col = sample)

# Here we just stack the two data.frame on top of each other (i.e. bind the rows together)
all_samples <- bind_rows(
  normal_samples,
  exponential_samples
) 

# Create QQ-plots
ggplot(all_samples,
       aes(sample = sample)) + 
  geom_qq() +
  facet_wrap(~i) +
  geom_abline(aes(slope = sd(sample), intercept = mean(sample)))

## To reveal the truth behind the i's, run this line:
all_samples %>% select(i, distribution) %>% unique()
```
