---
title: "Untitled"
author: "Ralph Trane"
date: "2/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

Import libraries:

```{r}
library(tidyverse) 
library(distributions3)
```

Create two random variables -- one Normal ($X$) and one Exponential ($Y$):

```{r}
X <- Normal(mu = 1, sigma = 1) 
Y <- Exponential(rate = 1)
```

Set the sample size.

```{r}
sample_size <- 100
```

Create a vector of i's to use. This will help us with the plot later. We don't want to simply use 1,2,...,16, since we would then know that the first 8 are normal, the last 8 exponential. We want to scrample (or permute) the numbers.

```{r}
is <- sample(1:16)
```

This next bit is a bit complicated, but do not worry -- you are not expected to know exactly what's going on. I suggest you run it one step at a time by highlighting and executing. 

* The first part gets a `data.frame` (i.e. data set) started with a column of 8 numbers
* we then `mutate` to create two new columns:
    - `distribution` will just tell us what variable have in this row
    - `sample`: the function `map` runs the function `random(X, sample_size)` for each value of `i`. I.e. we create 8 samples of size `sample_size` by drawing independently from a Normal(1,1). Each of the entries in this new columns will be a vector of numbers.
    - `unnest_longer` takes that new column of vectors and makes "unfolds" it. 
* in the end, we have a data.frame with three columns:
    - `i`: the sample number
    - `distribution` tells us that this is from a normal distribution
    - `sample` are the draws 

We do the same using the exponential distribution instead of the normal.


```{r}
normal_samples <- data.frame(i = is[1:8]) %>%
  mutate(distribution = "Normal", 
         sample = map(.x = i, .f = ~random(X, sample_size))) %>% 
  unnest_longer(col = sample) 

exponential_samples <- data.frame(i = is[9:16]) %>% 
  mutate(distribution = "Exponential",
         sample = map(i, ~random(Y, sample_size))) %>% 
  unnest_longer(col = sample)
```

To be able to create QQ-plots for all 16 samples at once, we combine the two `data.frame`s into one. We stack them on top of each other using `bind_rows`:

```{r}
all_samples <- bind_rows( 
  normal_samples, 
  exponential_samples
)
```

Finally, we create QQ-plots

```{r}
ggplot(all_samples,
       aes(sample = sample)) +
  geom_qq() +
  facet_wrap(~i) +
  geom_abline(aes(slope = sd(sample), intercept = mean(sample)))
```

To reveal the truth behind the i's, run this line:

```{r}
all_samples %>% select(i, distribution) %>% unique() %>% arrange(distribution, i)
```

