---
title: "Discussion 4 - Solution"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Consider a large population which has true mean $\mu$ and true standard deviation $\sigma$. We take a sample of size 3 from this population, thinking of the sample as the RVs $X_1, X_2, X_3$ where $X_i$  can be considered iid (independent identically distributed). We are interested in estimating $\mu$.

    a. Consider the estimator $\hat{\mu_1}=X_1 +X_2 -X_3$. What is the mean of this estimator? \newline
       **Solution:** $E(\hat{\mu_1}) = \mu$

    b. Find the variance of $\hat{\mu_1}$. \newline
       **Solution:** $\text{Var}(\hat{\mu_1}) = 3 \cdot \sigma^2$

    c. Consider the estimator $\hat{\mu_2}=\frac{X_1 + X_2 + X_3}{3}$. What is the mean of this estimator? \newline
       **Solution:** $E(\hat{\mu_2}) = \mu$

    d. Find the variance of $\hat{\mu_2}$.  
       **Solution:** $\text{Var}(\hat{\mu_2}) = \frac{\sigma^2}{3}$
    
    e. Now, consider the estimator $\hat{\mu_3}=\frac{X_1 +2X_2 +3X_3}{6}$. What is the mean of this estimator?  
       **Solution:** $E(\hat{\mu_3}) = \mu$

    f. Find the variance of $\hat{\mu_3}$.  
       **Solution:** $\text{Var}(\hat{\mu_3}) = \frac{7}{18}\sigma^2$
    
    g. Which of these three estimators is preferable? Why?  
       **Solution:** all unbiased, so on average, all will find the truth. However, $\mu_2$ has the smallest variance.

2. A packing plant fills bags with cement. The weight X kg of a bag of cement can be modeled by a normal distribution with mean 50kg and standard deviation 0.7kg.
	
	  a. Find $P(X > 51)$. Draw a sketch that visually indicates what this probability is. (I.e. normal curve, annotated with mean, SD, cut-off, shaded area)  
       **Solution:**

    ```{r}
    library(distributions3)
    library(tidyverse)
    X <- Normal(mu = 50, sigma = 0.7)
    
    1 - cdf(X, 51)
    ```
    
    
    ```{r echo = FALSE}
    plot_pdf(X) + geom_vline(linetype = 'dashed', aes(xintercept = 50, color = "Mean")) + geom_auc(from = 51)
    ```
    
    b. Three bags are selected randomly. Find the probability that at least two weigh more than 51kg.  
	     **Solution:** Since the bags are independent and sampled from the same distribution, each has the same chance to weigh over the threshold. So this is asking about a binomial distribution. The probability of "success" (a bag weighing over 51kg) was found in part (a) as 0.077. So letting $M \sim$ Bin(3, 0.077), we want $P(M \ge 2) = 1 - P(M \le 1)$. Use the binomial pmf or R.  
       In `R`, we can do either:

    ```{r}
    p <- 1-cdf(X, 51)
    Y <- Binomial(size = 3, p = p)
    
    1- cdf(Y,1)
    ```
        
        or
    
    ```{r}
    pmf(Y, 2) + pmf(Y, 3)
    ```

3. Weights of female cats are well approximated by a normal distribution with mean 4.1kg and standard deviation of 0.6kg $X \sim N(4.1, 0.6^2)$.

    a. What proportion of female cats have weights between 3.7 and 4.4kg? Draw a sketch that visually indicates what this probability is. (I.e. normal curve, annotated with mean, SD, cut-off, shaded area)  
       **Solution:**

    ```{r}
    X <- Normal(mu = 4.1, sigma = 0.6)
    cdf(X, 4.4) - cdf(X, 3.7)
    ```
    
    ```{r echo = FALSE}
    plot_pdf(X) + geom_auc(from = 3.7, to = 4.4) + geom_vline(aes(xintercept = 4.1, color = "mean"))
    ```
    
    b. A certain female cat has a weight that is 0.5 standard deviations above the mean. What proportion of female cats are heavier than this one? Draw a sketch that visually indicates what this probability is. (I.e. normal curve, annotated with mean, SD, cut-off, shaded area)  
       **Solution**

    ```{r}
    1 - cdf(X, 4.1 + 0.5*0.6)
    ```


    ```{r echo = FALSE}
    plot_pdf(X) + geom_auc(from = 4.1 + 0.5*0.6) + geom_vline(aes(xintercept = 4.1, color = "mean"))
    ```
    c. How heavy is a female cat whose weight is on the 80th percentile? Draw a sketch that visually indicates what value we are looking for. (I.e. normal curve, annotated with mean, SD, cut-off, shaded area)  
       **Solution**

    ```{r}
    quantile(X, 0.8)
    ```
       
    ```{r echo = FALSE}
    plot_pdf(X) + geom_vline(aes(xintercept = quantile(X, 0.8), color = "80th percentile"))
    ```
    d. A female cat is chosen at random. What is the probability that she weighs more than 4.5 kg? Draw a sketch that visually indicates what this probability is. (I.e. normal curve, annotated with mean, SD, cut-off, shaded area)  
       **Solution**

    ```{r}
    1-cdf(X, 4.5)
    ```

    ```{r echo = FALSE}
    plot_pdf(X) + geom_auc(from = 4.5)
    ```



4. Take a look at the `R`-code below. [**Note**: everything after a pound symbol (\#) is a comment, and NOT part of the `R` code.]

    a. Walk through it, one line at a time, and try to make sense of it. 
    
    b. Run all the code, **EXCEPT FOR THE LAST LINE**. This should generate a bunch of QQ-plots. Half of them are from a normal distribution, the other half an exponential distribution. Can you tell which correspond to samples from the normal distribution, and which correspond to samples from the exponential distribution?  
       **Solution:** I can maybe point out a few of the samples that come from the exponential distribution, but definitely not all.

    c. Repeat with `sample_size <- 100`. Does this change anything?  
       **Solution:** Now I can point out all the exponentials.

    d. What does this tell you about the use of QQ-plots to determine normality of data?  
       **Solution:** It sucks for small sample sizes.

```{r eval = T}
library(tidyverse)
library(distributions3)

## Create two random variables -- one Normal and one Exponential
X <- Normal(mu = 1, sigma = 1)
Y <- Exponential(rate = 1)

sample_size <- 10

## Create a vector of i's to use for scrambling
is <- sample(1:16)

normal_samples <- data.frame(i = is[1:8]) %>% 
  mutate(distribution = "Normal",                  # Create column that just says "Normal". 
                                                   # For book keeping.
         sample = map(i, random,                   # map runs the function "random" with
                      d = X, n = sample_size)) %>% # arguments d = X, and n = sample_size 
                                                   # for each i. I.e. it creates 8 random 
                                                   # samples of size sample_size from the 
                                                   # distribution X.
                                                   # The result is this weird list thing...
  unnest_longer(col = sample) # This get's rid of the list and simply gives us a 
                              # data.frame that we can work with...

exponential_samples <- data.frame(i = is[9:16]) %>% 
  mutate(distribution = "Exponential", 
         sample = map(i, random, d = Y, n = sample_size)) %>% 
  unnest_longer(col = sample)

# Here we just stack the two data.frame on top of each other (i.e. bind the rows together)
all_samples <- bind_rows(
  normal_samples,
  exponential_samples
) 

# Create QQ-plots
ggplot(all_samples,
       aes(sample = sample)) + 
  geom_qq() +
  facet_wrap(~i) +
  geom_abline(aes(slope = sd(sample), intercept = mean(sample)))

## To reveal the truth behind the i's, run this line:
all_samples %>% select(i, distribution) %>% unique() %>% arrange(distribution, i)
```



