---
title: "Lecture 3: Descriptive Statistics"
subtitle: "STAT 324"
author: "Ralph Trane"
institute: "University of Wisconsin--Madison<br><br>"
date: "Spring 2020"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [../css/uwmadison.css, default-fonts, ../css/extra-classes.css]
    lib_dir: libs
    nature:
      titleSlideClass: [center, top, .title-slide]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ration: '16:9'
      navigation:
        scroll: false
---

```{r include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, dpi = 300, fig.align = "center")
```


# Objectives

1. "Previously on STAT 324..."

2. Descriptive Statistics
    - Types of data/variables
    
    - Numerical summaries
    
    - Graphical summaries

---

# Previously on STAT 324...

* The general setup

--
* Three parts of this class:

    1. Descriptive Statistics

--
    2. Probability

--
    3. Inferential Statistics

---
layout: true

# Descriptive Statistics

---

* What: the art of describing data with few important measures ('summary statistics')

* Why: 
    - know your population!!
    - explore your data

* How: try to get an idea of the distributions of variables included 
    - what's a distribution?!
    - what's a variable?!

???

* What: when doing big studies, you collect lots of data. Take the essense of the data, summarize it in a few summary statistics, present it to the audience.

* Why: 
    - the sample is supposed to be representative of the population. Describe the sample to know what population you MIGHT be able to generalize to
    - often find things -- data entry errors, potential confounding variables, potential associations

* How: use numbers and especially figures to describe the distribution
    - what's a distribution? Next week
    - what's a variable -- today.

---

* Discrete data
    - categorical
        - no natural ordering
        - examples: sex, race, blood type, political orientation, etc. 
    - ordinal
        - naturally ordered
        - educational level, age groups, disease severity scales, counts, etc.
    - summarized by 
        - frequency counts
        - relative frequencies

--

* Continuous data
    - numerical
        - examples: age, height, weight, BMI, proportions, etc.
    - infinite (uncountable, actually...) number of potential values
    - summarized by
        - location measures
        - spread/variation measures
        - shape

---
layout: true

# Descriptive Statistics

**Data example: STAR WARS**

---

First, some `R` setup:

```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(DT); library(kableExtra)

# Change theme for plots
theme_set(theme_bw()) 

# Change color scheme for plots
scale_color_continuous <- scale_color_viridis_c
scale_color_discrete <- scale_color_viridis_d

scale_fill_continuous <- scale_fill_viridis_c
scale_fill_discrete <- scale_fill_viridis_d
```

---

Let's load some data, and take a look:

```{r starwarsdata}
starwars <- read_csv("../../csv_data/starwars.csv")

## datatable and formatStyle are from the DT package
datatable(starwars, options = list(pageLength = 4, lengthChange = FALSE,
                                   scrollX = TRUE, dom = "tp")) %>% 
  formatStyle(columns = 1:ncol(starwars), fontSize = "12pt")
```

---

How many films have the characters been in? 

Ordinal variable. Summarized using frequency counts, and relative frequencies.

Numerical summary:

.pull-left[
```{r starwars1, eval = FALSE}
n_films <- starwars %>% 
  group_by(n_films) %>% 
  summarize(frequency = n()) %>%
  mutate(relative = frequency / sum(frequency))

kable(n_films, 
      format = "html") %>% 
  kable_styling(font_size = 12) 
# kable_styling is from the kableExtra package.
```
]

.pull-right[

```{r starwars1-out, ref.label="starwars1", echo = FALSE, warning = FALSE, message = FALSE}
```

]

---

How many films have the characters been in? 

Ordinal variable. Summarized using frequency counts, and relative frequencies.

Graphical summary:

.pull-left[
```{r starwars2, eval = FALSE}
ggplot(data = starwars,
       aes(x = n_films)) +
  geom_bar() +
  labs(x = "Number of films", 
       y = "Frequencies")
```
]

.pull-right[
```{r starwars2-out, ref.label="starwars2", echo = FALSE, warning = FALSE, message = FALSE, fig.width=2, fig.height=2, dpi = 300, out.width = "100%"}
```
]

---

How many films have the characters been in? 

Ordinal variable. Summarized using frequency counts, and relative frequencies.

Graphical summary:

.pull-left[
```{r starwars3, eval = FALSE}
ggplot(data = starwars,
       aes(x = n_films)) +
  geom_bar(aes(y = ..count../sum(..count..))) +
  labs(x = "Number of films", 
       y = "Relative frequencies")
```
]

.pull-right[
```{r starwars3-out, ref.label="starwars3", echo = FALSE, warning = FALSE, message = FALSE, fig.width=2, fig.height=2, dpi = 300, out.width = "100%"}
```
]

---

When would we use relative frequencies rather than frequencies?
--
 When comparing groups.

.pull-left[
```{r compare-groups, eval = FALSE}
n_films_gender <- starwars %>% 
  group_by(gender, n_films) %>% 
  summarize(frequencies = n()) %>% 
  mutate(relative = frequencies / sum(frequencies))

kable(n_films_gender,
             format = "html") %>% 
  kable_styling(font_size = 10)
```
]

.pull-right[
```{r ref.label="compare-groups", echo = FALSE}
```
]
---

When would we use relative frequencies rather than frequencies? When comparing groups.

```{r compare-groups-fig, fig.height = 2.75, dpi = 300}
ggplot(data = n_films_gender,
       aes(x = n_films, y = frequencies, fill = gender)) + 
  geom_bar(stat = "identity") + 
  facet_grid(~gender) + 
  labs(x = "Number of films")
```

---

When would we use relative frequencies rather than frequencies? When comparing groups.

```{r fig.height = 2.75, dpi = 300}
ggplot(data = n_films_gender,
       aes(x = n_films, y = relative, fill = gender)) + 
  geom_bar(stat = "identity") + 
  facet_grid(~gender) + 
  labs(x = "Number of films")
```


---
layout: true

# Descriptive Statistics

---

For continuous data, measures of location and spread

Measures of location:

* mean/average: $\bar{x} = \frac{1}{n}\left(x_1 + x_2 + ... + x_n\right) = \frac{1}{n} \sum_{i=1}^n x_i$
* median: the observation in the middle of the data. 
* minimum and maximum: the smallest and largest observations, respectively

--

```{r echo = FALSE}
silly_data <- c(1, 5, 2, -3, 7, -12, 0)
```

Simple example data: $`r paste(silly_data, collapse = ", ")`$. 

Mean:
--
 `r mean(silly_data)`

Median: 
--
 `r median(silly_data)`

Minimum and maximum: 
--
 `r min(silly_data)` and `r max(silly_data)`


---

Measures of spread:

* variance: $s^2 = \frac{1}{n-1} \sum_{i=1}^n (x_i - \bar{x})^2$
* standard deviation: $s = \sqrt{s^2}$
* range: $\text{max} - \text{min}$
* percentiles (sometimes referred to as quantiles): 
    
> The *p'th percentile* of a set of measurements is the one that has $p \%$ of the data below it.

* Inter Quartile Range (IQR): 75th percentile - 25th percentile
    - 25th percentile = first quartile = median of 1st half including median
    - median = second quartile
    - 75th percentile = third quartile = median of 2nd half including median

--

Simple example data: $`r paste(silly_data, collapse = ", ")`$. 

Variance:
--
 `r var(silly_data)`. Standard deviation:
--
 `r sd(silly_data)`. Range:
--
 $`r max(silly_data) - min(silly_data)`$. <br>
1st and 3rd quartiles (i.e. 25th and 75th percentiles):
--
 `r quantile(silly_data, p = c(0.25, 0.75))`. IQR:
--
 `r IQR(silly_data)`

Why don't we use these summaries for discrete data?
--
 Doesn't make sense for categorical data. Could maybe use these in some cases when working with ordinal data, although interpretation might be hard.
---
layout: true

# Descriptive Statistics

**Data example: STAR WARS**

---

```{r many-summaries, eval = FALSE}
starwars %>% 
  filter(!is.na(height), gender %in% c('male', 'female')) %>% 
  group_by(gender) %>% 
  summarize(n = n(),
            mean = mean(height),
            var = var(height),
            sd = sd(height),
            median = median(height),
            Q1 = quantile(height, p = 0.25),
            Q2 = quantile(height, p = 0.50),
            Q3 = quantile(height, p = 0.75),
            IQR = IQR(height), 
            min = min(height),
            max = max(height),
            range = max - min) %>% 
  kable(format = "markdown", digits = 2)
```

---

```{r ref.label="many-summaries", echo = FALSE}
```

---

```{r}
starwars %>% 
  filter(!is.na(height), gender %in% c('male', 'female')) %>% 
  group_by(gender) %>% 
  summarize(n = n(),
            '10th percentile' = quantile(height, p = 0.1),
            '57th percentile' = quantile(height, p = 0.57),
            '82nd percentile' = quantile(height, p = 0.82)) %>% 
  kable(format = "markdown", digits = 2)
```



---

Visually, continous data can be summarized using a *histogram*. 

```{r fig.width = 5, fig.height = 2, out.height = "45%", out.width = "90%"}
ggplot(data = starwars,
       aes(x = height)) + 
  geom_histogram(binwidth = 10, boundary = 60,
                 na.rm = TRUE, color = 'black')
```

How does mean and variance relate to a histogram? 

---

Proportion of an area is the proportion of data in the corresponding interval: proportion of characters with height less than 100 $= \frac{7 \cdot 10}{`r nrow(filter(starwars, !is.na(height)))` \cdot 10} \approx 0.086$. 

This will be **very** important to us later. 

```{r echo = FALSE, fig.width = 5, fig.height = 3, out.height = "50%", out.width = "90%"}
plotly::ggplotly(
  ggplot(data = starwars,
         aes(x = height)) + 
    geom_histogram(binwidth = 10, boundary = 60, aes(fill = height <= 100),
                   na.rm = TRUE, color = 'black') +
    scale_fill_manual(values = c("grey40", "red")) + 
    theme(legend.position = "none"), tooltip = "count"
)
```

---

```{r echo = FALSE, fig.width = 5, fig.height = 3, out.width = "100%"}
ggplot(data = starwars %>% filter(gender %in% c('male', 'female')),
       aes(x = height)) +
  geom_histogram(binwidth = 10, boundary = 60, na.rm = T) +
  geom_vline(data = starwars %>% 
               filter(gender %in% c('male', 'female')) %>% 
               group_by(gender) %>% 
               summarize(mean = mean(height, na.rm = T),
                         median = median(height, na.rm = T),
                         `90th percentile` = stats::quantile(height, p = 0.9, na.rm = T)) %>% 
               pivot_longer(cols = c(mean, median, `90th percentile`), names_to = 'stat', values_to = 'value'),
             aes(xintercept = value, color = stat)) +
  facet_grid(gender ~ .) + 
  scale_color_discrete("") +
  theme(legend.position = "top")
```

---

Histograms can also be used to depict *relative frequencies*. To do so, you simply divide the counts by the total number of observations.

```{r echo = FALSE, fig.width = 5, fig.height = 2, out.height = "50%", out.width = "90%"}
ggplot(data = starwars,
       aes(x = height)) + 
  geom_histogram(binwidth = 10, boundary = 60,
                 na.rm = TRUE, color = 'black',
                 aes(y = ..count../sum(..count..)))
```

---

Another useful figure for numerical data: the boxplot (also known as box-and-whisker plot, or box-and-whisker diagram).

The key to decipher a box plot:

.pull-left[
Outlier if 

$\text{obs} < \text{Q1} - 1.5\cdot \text{IQR}$ 

or 

$\text{obs} > \text{Q3} + 1.5\cdot \text{IQR}$


25% of the data are above the box

50% of the data are in the the box

25% of the data are below the box


]

.pull-right[
```{r echo = FALSE}
ggplot(data = data.frame(y = c(1,2,5,10,10,10,11,13,13,13,15,15,15,15,16,20,24)),
       aes(x = 1, y = y)) + 
  geom_boxplot(width = 0.5) + 
  geom_text(data = data.frame(y = c(1.5,5,10,13,15,20,24),
                               labels = c("outliers", "minimum\n(excluding outliers)",
                                          "Q1", "median", "Q3", 
                                          "maximum\n(excluding outliers)", "outlier")),
            aes(x = 1.75, label = labels),
            hjust = 0, size = 4.75) +
  geom_segment(data = data.frame(y = c(1.5,1.5,5,10,13,15,20,24),
                                 yend = c(1,2,5,10,13,15,20,24)),
               aes(x = 1.65, xend = c(1.1, 1.1, 1.1, 1.35, 1.35, 1.35, 1.1, 1.1),
                   y = y, yend = yend),
               arrow = arrow(length = unit(0.03, "npc")),
               color = 'darkred') +
  theme_minimal() +
  labs(x = '', y = '') +
  scale_x_continuous("", limits = c(0,2.2)) +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        plot.margin=grid::unit(c(0,0,0,0), "mm"))
```
]

---


```{r fig.height = 3, fig.width = 3, out.height = "300px", fig.align="center"}
boxplots <- ggplot(data = starwars %>% filter(gender %in% c('male', 'female')),
       aes(x = gender, y = height)) +
  geom_boxplot()

boxplots
```

---

My personal favorite when data set not too large: boxplot + points!

```{r boxplot-plus-points, fig.height = 3, fig.width = 3, out.height = "300px", fig.align="center"}
boxplots +
  geom_point(aes(color = gender), alpha = 0.5,
             position = position_jitter(width = 0.2)) +
  guides(color = "none")
```

---
layout: true

# Descriptive Statistics

**Continuous Data: shape**

---


So far, only talked about *location* and *spread* of data. 

--

However, this does not provide the full picture:

```{r 4_dists, echo = FALSE, out.height = "350px", fig.width = 8, fig.height = 4}
mu <- 2
var <- mu^2

exp_data <- rexp(1000, rate = 1/mu)
norm_data <- rnorm(1000, mean = mu, sd = sqrt(var))
beta_data <- rbeta(1000, shape1 = 0.9, shape2 = 0.25)
for_bimodal <- rbernoulli(1000, p = 0.75)
bimodal_data <- for_bimodal*rnorm(1000, -3) + (1-for_bimodal)*rnorm(1000, 3) #runif(1000, min = 0, max = 6.928203)

dat <- data.frame(data = c(exp_data, norm_data, beta_data, bimodal_data),
                  dist = rep(1:4, each = 1000)) %>% 
  group_by(dist) %>% 
  mutate(data = (data - mean(data))/sd(data)*mu + mu,
         label = paste0("Mean = ", mean(data), "; Variance = ", var(data)))

dist_hists <- ggplot(data = dat,
       aes(x = data)) + 
  geom_histogram(bins = 50) +
  theme(legend.position = "top")

dist_hists + 
  facet_wrap(~dist, scales = "free", nrow = 1,
             labeller = labeller(dist = rep(unique(dat$label), 4) %>% setNames(1:4)))
```

---

The *shape* of the data is generally described as either *symmetrical* or *non-symmetrical*, and if non-symmetrical as either *right skewed* or *left-skewed*. 

When symmetrical, mean = median. Right skewed, mean > median. Left skewed, mean < median.

```{r echo = FALSE, out.height = "350px", fig.width = 8, fig.height = 4}
dist_hists + 
  geom_vline(data = dat %>% 
               group_by(dist) %>% 
               summarize(mean = mean(data),
                         median = median(data)) %>% 
               pivot_longer(cols = c(mean, median), 
                            names_to = "stat"),
             aes(xintercept = value, color = stat)) +
  facet_wrap(~dist, scales = "free", nrow = 1,
             labeller = labeller(dist = c("Right skewed", "Symmetrical", "Left skewed", "Asymmetrical/bimodal") %>% setNames(1:4)))
```

---

Boxplots don't give us the entire shape, but they let us determine if the data is symmetrical: 

.pull-left[
For symmetrical data

* median is in the middle of the data.
* median = mean. 

Right skewed data => median closer to bottom of box.

Left skewed data => median closer to top of box.
]

.pull-right[
```{r out.height = "300px", echo = FALSE, fig.width = 2, fig.height = 2}
ggplot(data = dat,
       aes(x = as.character(dist), y = data)) + 
  geom_boxplot() + 
  labs(x = "") + 
  theme(legend.position = "top")
```
]

