---
title: "Cool stuff"
author: "Ralph Trane"
output: html_document
---

Load packages:

```{r}
library(tidyverse)
```


## Historic CO2

Contains CO2 data:

* Carbon dioxide concentration in ppm by volume
* year
* source

```{r}
historic_co2 <- read_csv("data/historice_co2.csv")

co2_plot <- ggplot(data = historic_co2, 
                   aes(x = year, y = co2, color = source)) + 
  geom_line()

co2_plot
```

Using a different package (`plotly`), we can make this interactive with one (!!) line of code!

```{r}
library(plotly)
ggplotly(co2_plot)
```

## Murders



```{r}
murders <- read_csv("data/murders.csv")
```

Contains

* US state
* abbreviation
* region
* population in 2010
* number of gun murders in 2010

Overall average

```{r}
murders %>% 
  summarize(mean_total = mean(total))
```

Any state differences? 

```{r}
murders %>% 
  group_by(state) %>% 
  summarize(mean_total = mean(total))
```

Any region differences?

```{r}
murders %>% 
  group_by(region) %>% 
  summarize(mean_total = mean(total))
```

Should probably do per capita:

```{r}
murders %>% 
  mutate(n_per_1000 = total / population * 1000) %>% 
  group_by(region) %>% 
  summarize(mean_total = mean(n_per_1000))
```

South seems higher. But any difference between Northeast and North Central? 

* Briefly mention sampling error: "if we sampled different year, see same?"
* Anything else that hints to the need for more analysis

Let's create a pretty map. Need some state data first:

```{r}
states_map <- map_data("state")
```

Create the map, and save it for later use:

```{r}
murders_by_state <- ggplot(murders, aes(map_id = state)) +
  geom_map(aes(fill = total / population * 1000), 
           map = states_map) +
  coord_map() +
  expand_limits(x = states_map$long, y = states_map$lat)
```

Show map

```{r}
murders_by_state
```


Interactive:

```{r}
ggplotly(murders_by_state)
```

## Research Funding

Read data:

```{r}
research_funding_rates <- read_csv("data/research_funding_rates.csv")
```

Overall success rates:

```{r}
research_funding_rates %>% 
  group_by(gender) %>% 
  summarize(total_applications = sum(applications),
            total_awards = sum(awards)) %>% 
  mutate(success_rate = total_awards / total_applications)
```

By discipline:

```{r}
research_funding_rates %>% 
  mutate(success_rate = awards / applications)
```

Bar chart:

```{r}
success_rates <- research_funding_rates %>% 
  mutate(success_rate = awards / applications)

ggplot(data = success_rates,
       aes(x = discipline, y = success_rate, fill = gender)) + 
  geom_bar(stat = 'identity', position = position_dodge()) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))
```

## Interactive Maps (Google style)

```{r}
library(leaflet)

leaflet() %>% addTiles()
```


```{r}
leaflet() %>% 
  addProviderTiles(provider = "Esri.WorldImagery",
                   options = tileOptions(minZoom = 2,
                                         noWrap = TRUE)) %>% 
  addPopups(lng = c(-89.405032, -89.407260, 11.324524), 
            lat = c(43.074805, 43.074241, 55.257877), 
            popup = c("We're here.", "MSC", "Denmark!"),
            options = popupOptions(closeButton = FALSE,
                                   closeOnClick = FALSE)) %>% 
  setView(lng = -89.405032, lat = 43.074805, zoom = 20)
```

