---
title: "Homework 0"
author: "Ralph Trane"
date: "1/21/2020"
output: html_document
---

The following is the product of great collaboration with Nathanial. 

First we import packages we will need.

```{r}
library(tidyverse)
```

We read the data into R. 

```{r}
measles <- read_csv("measles.csv") 
```

# Question 1

We have 6 variables in the data set. They are disease, state, year, weeks_reporting, count, and population. 

# Question 2 

The data have 3876 observations. 

# Question 3

Using `filter`, we can get just the data for Wisconsin in 1993. 

```{r}
filter(measles, state == "Wisconsin", year == 1993)
```

# Question 4

To find the mean rate by state we first create a new varible called `rate` (using `mutate`), then `group_by` state before summarizing the data. We use `arrange` to find the smallest and largest means. 

```{r}
measles %>% 
  mutate(rate = count / population) %>% 
  group_by(state) %>% 
  summarize(mean = mean(rate)) %>% 
  arrange(mean)
```

The state with the overall lowest mean rate is Mississippi. 

```{r}
measles %>% 
  mutate(rate = count / population) %>% 
  group_by(state) %>% 
  summarize(mean = mean(rate)) %>% 
  arrange(desc(mean))
```

The state with the overall largest mean rate is Wisconsin. 


# Question 5

Data is missing for Alaska and Hawaii, so we remove them. The we create the rate variable, `group_by` year, and summarize to get the mean rate per year. The result is saved to an object called `yearly_mean_rate`. 

```{r}
yearly_mean_rate <- measles %>% 
  filter(state != "Alaska", state != "Hawaii") %>% 
  mutate(rate = count / population) %>% 
  group_by(year) %>% 
  summarize(mean = mean(rate))
```

To actually show the result:

```{r}
yearly_mean_rate
```

Create a plot:

```{r}
ggplot(data = yearly_mean_rate, 
       aes(x = year, y = mean)) + 
  geom_line()
```

From the plot, it seems like something significant happened around 1960.

# Question 6

Add a vertical line using `geom_vline`.

```{r}
ggplot(data = yearly_mean_rate, 
       aes(x = year, y = mean)) + 
  geom_line() + 
  geom_vline(xintercept = 1963)
```

Just for Wisconsin and Alabama.

```{r}
ggplot(data = filter(measles, state %in% c("Wisconsin", "Alabama")),
       aes(x = year, y = count / population, color = state)) +
  geom_line()
```

# Bonus

Pretty plot using `geom_tile`, and make the vertical line red. 

```{r}
ggplot(data = measles, 
       aes(x = year, y = state, fill = count / population)) + 
  geom_tile(color = "grey") +
  geom_vline(xintercept = 1963, color = "red")
```

